---
title: "Emergy — Odum's quality-of-energy lens"
pagetitle: "emburden.org — Emergy"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", out.width="100%",
                      fig.width=10, fig.height=6, dpi=110)
suppressPackageStartupMessages({library(dplyr); library(ggplot2); library(sf)})
DATA_DIR <- "../docs/global_analysis_data"
# Bundled brand palette (matches emburdensynR/brand_palette.R:emrgi_brand_colors)
BR <- c(g = "#228b22", gd = "#176b17", gdd = "#0f4d0f", gold = "#b5791b",
        ink = "#1d2a1d", mut = "#566656", line = "#d6e4d6", bg = "#f3f8f3",
        charge = "#c0392b", warn = "#8a5a16")
source("R/counts.R")
em <- readRDS(file.path(DATA_DIR, "emergy_metrics.rds"))
em_c <- em$country
em_cell <- em$cell
world <- tryCatch(readRDS(file.path(DATA_DIR, "world_sf.rds")), error=function(e) NULL)
```

# Emergy — a different currency for energy analysis

Climate-justice accounting uses **CO~2~ emissions** as the currency.
Odum's emergy framework uses **solar-energy equivalent embodied per
joule delivered** (measured in solar emjoules, sej). Same physical
kilowatt-hour at the socket can carry wildly different prior emergy
investment depending on the fuel that generated it.

<div class="callout">
<div class="callout-title">Why care about a second currency</div>
<p>Two grids that deliver identical kWh per capita can differ
<b>5-fold or more</b> in the electricity-basis emergy they concentrate
into each joule (Bundle-A-corrected range: `r sprintf("%.1fx", max(em_c$transformity_sej_per_J, na.rm=TRUE) / min(em_c$transformity_sej_per_J, na.rm=TRUE))` across
`r nrow(em_c[!is.na(em_c$transformity_sej_per_J), ])` countries). That
matters for questions climate-justice accounting cannot answer:
"how much biophysical infrastructure did this energy require to
produce?" and "how sustainably is a country converting environmental
subsidy into human welfare?"</p>
</div>

## The transformity spectrum

Every fuel carries a canonical **transformity** — the sej required per
joule of the output. Odum 1996's table + Brown & Ulgiati 2004's
revised baseline give us:

| Fuel | Transformity (sej/J) | Where in the emergy hierarchy |
|---|---:|---|
| Wind (kinetic) | 1,500 | Low — little prior emergy invested per J |
| Solar-PV (electricity) | 3,400 | Low — panels themselves are direct-solar |
| Bioenergy (wood, dung) | 25,000 | Medium — biosphere × sunlight × years |
| Coal (heat) | 40,000 | Medium — plants × millions of years × geologic pressure |
| Natural gas (heat) | 48,000 | Medium |
| Oil (heat) | 54,000 | Medium-high |
| Hydro (electricity) | 80,000 | High — watershed × millennia + concentrator dam |
| Nuclear (electricity) | 200,000 | Highest — fuel cycle + massive infrastructure |

Notice something climate-accounting hides: **hydro and nuclear are
both "clean" in CO~2~ terms but sit at the highest transformity tier.**
They represent massive prior-emergy investment concentrated into each
joule. Wind and solar deliver similar joules with far less prior
emergy locked in.

## What our data reveals across `r nrow(em_c)` countries

```{r tr-range, results="asis"}
em_c_ok <- em_c[!is.na(em_c$transformity_sej_per_J), ]
tr_max <- max(em_c_ok$transformity_sej_per_J)
tr_min <- min(em_c_ok$transformity_sej_per_J)
tr_ratio <- tr_max / tr_min
top5 <- em_c_ok[order(-em_c_ok$transformity_sej_per_J), ][1:5, ]
bot5 <- em_c_ok[order( em_c_ok$transformity_sej_per_J), ][1:5, ]
fmt_k <- function(x) sprintf("%.0fk", x / 1000)
cat(sprintf(
"**Grid transformity ranges %.1fx across the world's fuel mixes** (%.0f-%.0f sej/J, on a per-J-electricity-delivered basis after Bundle-A efficiency correction). From `emergy_metrics.rds`:\n\n",
tr_ratio, tr_min, tr_max))
cat("- **Top-5 highest transformity** (nuclear-heavy grids): ",
    paste(sprintf("%s ~%s", top5$iso3, fmt_k(top5$transformity_sej_per_J)),
          collapse = ", "), " sej/J\n", sep = "")
cat("- **Bottom-5 lowest transformity** (wind + solar-dominant grids): ",
    paste(sprintf("%s ~%s", bot5$iso3, fmt_k(bot5$transformity_sej_per_J)),
          collapse = ", "), " sej/J\n", sep = "")
```

The **key insight** for household energy policy: physically identical
kilowatt-hours can carry very different embodied emergy depending on
the grid mix. Policy that treats them as interchangeable is missing
quality-adjusted throughput.

## Three findings unique to the emergy lens

<div class="card-grid">

<div class="card">
<div class="kicker">Finding 1</div>
<h3>Renewable ≠ low emergy</h3>
<p>Hydro-heavy grids (Norway, Paraguay, Ethiopia) carry <b>high</b>
transformity because dam infrastructure + watershed × millennia
concentrate huge emergy per J. Wind-heavy grids (Denmark) carry
<b>low</b> transformity — they harvest kinetic energy without a large
prior emergy investment. The "renewable" label conflates two Odum
categories. Emergy separates them.</p>
</div>

<div class="card">
<div class="kicker">Finding 2</div>
<h3>Empower per capita is a biophysical standard-of-living metric</h3>
<p>Country empower / population — sej/person/yr — correlates only
weakly with GDP per capita. Nuclear-heavy France scores much higher
per person than solar-heavy California at similar income. Suggests
GDP misses a real dimension of energy-based welfare.</p>
</div>

<div class="card">
<div class="kicker">Finding 3</div>
<h3>Household emergy-per-dollar arbitrage exposes hidden subsidies</h3>
<p>When a household's sej-per-dollar exceeds the national average,
they're receiving a subsidised bargain on high-emergy energy —
typical in nuclear-heavy grids with residential tariffs subsidised
below marginal cost. When it's <em>below</em>, they're paying a
premium for low-emergy energy — typical in biomass + diesel-dependent
low-access economies where transaction costs eat their expenditure.</p>
</div>

</div>

## Bringing it together — the Household Emergy Insecurity Index (HEII)

The three findings above each look at one axis. **Emergy's real power
is that it lets us combine physical adequacy and economic burden into
a single number** — because both convert to the same currency (sej):

<div class="callout">
<div class="callout-title">The unified formulation</div>
<pre>
physical_gap_sej = max(0, required_empower  − delivered_empower)
economic_gap_sej = max(0, paid_empower      − affordable_empower)

heii_sej      = physical_gap_sej + economic_gap_sej          (sej/yr)
heii_norm     = heii_sej / income_empower                    (fraction)
heii_esi      = (delivered/required) / (paid/affordable)     (Odum-canonical)
heii_product  = adequacy × (1 − unaffordability)             (∈ [0, 1])
</pre>
</div>

All three unified forms are computed by
[`household_emergy_insecurity()`](https://github.com/ericscheier/emburdensynth/blob/global/R/emergy.R)
in the emburden R module. The "affordable" threshold uses the national
em/\$ rate at a 10% burden anchor by default; other conventions
(6%, manual, national) are one-argument switches.

### What the HEII world reveals

```{r heii-summary, results="asis"}
em <- readRDS(file.path(DATA_DIR, "emergy_metrics.rds"))
heii <- em$country[!is.na(em$country$heii_norm), ]
top5 <- heii$iso3[order(-heii$heii_norm)][1:5]
bot5 <- heii$iso3[order( heii$heii_norm)][1:5]
cat(sprintf("Across the **%d countries** with complete kWh + spend + income + fuel-mix + GDP data, HEII_norm ranges 0 to %.0f. The five most insecure are `%s` — all Sub-Saharan Africa or Afghanistan, all bearing BOTH physical rationing (delivered kWh far below the physics baseline) AND economic overreach (energy expenditure exceeding 10%% of income emergy). The five most secure — `%s` — are petro-states + subsidised-tariff economies where households are both fully served and pay below the affordability threshold.\n",
  nrow(heii), max(heii$heii_norm),
  paste(top5, collapse=", "), paste(bot5, collapse=", ")))
```

### Bundle B — biomass cooking counted (2026-09)

Pre-Bundle-B, HEII was **grid-electricity only**. That
misrepresented ~2.4B people who cook primarily with biomass — the WHO
Household Energy Database says `r sum(heii$biomass_share > 0.5, na.rm = TRUE)`
of our HEII countries have >50% of households using non-clean cooking
fuels. Their firewood + charcoal + kerosene deliver real emergy that
was invisible to grid-only accounting.

`household_emergy_insecurity()` now accepts an `hh_fuels` argument
that credits direct-combustion fuels via
`household_fuel_transformity_table()` (wood 25k sej/J, charcoal 45k,
kerosene 54k, LPG 48k, piped-gas 48k, dung 20k). The builder script
weights an 8 GJ/yr wood-cookstove household by each country's WHO
`clean_fuel_access_pct` residual. Effect on the ranking:

```{r heii-biomass-shift, results="asis"}
if ("heii_norm_grid_only" %in% names(heii)) {
  shift <- heii[order(-heii$biomass_share), c("iso3", "biomass_share",
                                                "heii_norm_grid_only",
                                                "heii_norm")][1:5, ]
  cat("Top 5 biomass-cooking countries — HEII_norm change when biomass is counted:\n\n")
  cat("| iso3 | biomass share | HEII (grid only) | HEII (with biomass) |\n")
  cat("|---|---:|---:|---:|\n")
  for (i in seq_len(nrow(shift)))
    cat(sprintf("| %s | %.0f%% | %.1f | %.1f |\n",
                shift$iso3[i], 100 * shift$biomass_share[i],
                shift$heii_norm_grid_only[i], shift$heii_norm[i]))
}
```

The interpretation is important: these households are still
physically rationed on the electricity side (they can't run a fridge
or an efficient stove), but their emergy delivery is not zero. The
policy question shifts from "why do they have no energy" to "why is
what they get so low-quality" (open-fire wood at 10-30% burner
efficiency, PM2.5 exposure, deforestation externalities not priced).

## Time is now on the record — Bundle C

The library-layer functions have always accepted a `year` argument,
but the compiled cache used to be a snapshot at each country's
latest Ember year. `scripts/build_emergy_panel.R` (Bundle C) now
produces `emergy_metrics_panel.rds` — a `(iso3, year)` grid
covering **2000-2025** across ~213 countries. Trajectories over 25
years are readable directly:

```{r fig-transformity-trajectory, fig.height=4.8, fig.width=9, fig.cap="Grid transformity 2000-2024 for a selection of large economies. Germany's Energiewende drops transformity from ~136k to ~60k sej/J as renewables rise from 6% to 59%."}
panel_p <- tryCatch(readRDS(file.path(DATA_DIR, "emergy_metrics_panel.rds")),
                     error = function(e) NULL)
if (!is.null(panel_p)) {
  focus <- c("DEU", "USA", "CHN", "IND", "GBR", "FRA", "BRA", "JPN")
  pp <- panel_p$panel |>
    filter(iso3 %in% focus, year >= 2000, year <= 2024,
           is.finite(transformity_sej_per_J))
  ggplot(pp, aes(x = year, y = transformity_sej_per_J,
                  colour = iso3, group = iso3)) +
    geom_line(linewidth = 1.1) + geom_point(size = 1.3) +
    scale_y_continuous(labels = function(x) sprintf("%.0fk", x/1000)) +
    labs(x = NULL, y = "Grid transformity (sej/J, electricity-basis)",
         title = "Transformity 2000-2024 — 8 major economies") +
    theme_minimal(base_size = 11)
}
```

**What today's panel covers**: transformity, %R, ESI, empower per
capita, em/USD for every (country, year) with Ember coverage
(2000-2025). **What's still snapshot**: HEII needs per-year household
inputs (median_kwh_hh, spend, income) that we don't yet publish as
panels; the physics-required baseline uses a single climate normal;
retail prices are a 2022-2024 vintage that carries forward.
Follow-up work — see [`FUTURE_WORK.md`](https://github.com/ericscheier/emburdensynth/blob/global/FUTURE_WORK.md).

<div class="callout">
<div class="callout-title">Backward-compatible with the existing insecurity score</div>
<p>The Round-19 <code>country_unified_metrics()</code> function
(EJ-native) computed <code>insecurity_score = max(0, 1 −
actual_ej/required_ej)</code> — physical adequacy only, no economic
term. HEII is additive to that: it retains physical adequacy AND
introduces the economic axis in the same currency. Both are
preserved in the cache; <code>unified_comparison()</code> joins them
for cross-checking. See Layer M.7 in the
<a href="global_analysis.pdf">tech report</a> for the sign-agreement
diagnostic.</p>
</div>

## The math (short version)

```r
library(emburdensynth)

# Transformity of the fuel mix in a country-year (electricity-basis,
# post-Bundle-A: fossil transformities divided by their electric efficiency)
grid_transformity("NOR", 2023)   # hydro-heavy — moderate (~73k sej/J)
grid_transformity("FRA", 2023)   # nuclear-heavy — highest tier (~150k)
grid_transformity("DNK", 2023)   # wind + biomass — biomass lifts it (~33k)
grid_transformity("POL", 2023)   # coal-heavy — high (~88k, was ~35k pre-Bundle-A)

# Household empower — sej/yr delivered per household
household_empower("USA", 2020, kwh_per_year = 11000)  # sej per year

# Emergy Sustainability Index (Odum)
household_ESI("NOR", 2020)  # named vec: R_pct, NR_pct, ELR, EYR, ESI

# National empower / GDP
emergy_per_money("FRA", 2020)                    # sej per $
```

All functions ship in the [`emburden` R package](https://github.com/ericscheier/emburdensynth/blob/global/R/emergy.R).
The full metric table for `r nrow(em_c)` countries is at
[`docs/global_analysis_data/emergy_metrics.rds`](https://github.com/ericscheier/emburdensynth/tree/glob../docs/global_analysis_data).

<div class="callout warning">
<div class="callout-title">Additive, not replacement</div>
<p>Emergy accounting does not supersede CO<sub>2</sub> accounting. They
answer different questions. The climate-justice mismatch (67× per-capita
cumulative CO<sub>2</sub> disparity) still stands as a headline for who
pays what. Emergy adds a quality axis: <em>what kind</em> of energy
they pay for.</p>
</div>

## The six panels

The [technical report](global_analysis.pdf) (Layer M, pp 82–91) shows
six emergy figures:

1. Grid transformity world map (sej/J)
2. Country empower per capita world map (sej/person/yr)
3. ESI vs GDP-pc scatter (weak correlation)
4. Household em/USD arbitrage scatter
5. Cell-level %R world map (17,028 cells)
6. Cell-level emergy insecurity gap (sej/yr)

See also: [The Atlas](atlas.html) · [Methods](methods.html) ·
[Data provenance](data.html) · [Papers](papers.html) ·
[Code](code.html)

<footer>
<p><a href="/">← back to home</a> · <a href="access.html">access + affordability + QoS →</a></p>
</footer>
