---
title: "Emergy — Odum's quality-of-energy lens"
pagetitle: "emburden.org — Emergy"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", out.width="100%",
                      fig.width=10, fig.height=6, dpi=110)
suppressPackageStartupMessages({library(dplyr); library(ggplot2); library(sf)})
DATA_DIR <- "../docs/global_analysis_data"
# Bundled brand palette (matches emburdensynR/brand_palette.R:emrgi_brand_colors)
BR <- c(g = "#228b22", gd = "#176b17", gdd = "#0f4d0f", gold = "#b5791b",
        ink = "#1d2a1d", mut = "#566656", line = "#d6e4d6", bg = "#f3f8f3",
        charge = "#c0392b", warn = "#8a5a16")
source("R/counts.R")
em <- readRDS(file.path(DATA_DIR, "emergy_metrics.rds"))
em_c <- em$country
em_cell <- em$cell
world <- tryCatch(readRDS(file.path(DATA_DIR, "world_sf.rds")), error=function(e) NULL)
```

# Emergy — a different currency for energy analysis

Climate-justice accounting uses **CO~2~ emissions** as the currency.
Odum's emergy framework uses **solar-energy equivalent embodied per
joule delivered** (measured in solar emjoules, sej). Same physical
kilowatt-hour at the socket can carry wildly different prior emergy
investment depending on the fuel that generated it.

<div class="callout">
<div class="callout-title">Why care about a second currency</div>
<p>Two grids that deliver identical kWh per capita can differ
<b>5-fold or more</b> in the electricity-basis emergy they concentrate
into each joule (Bundle-A-corrected range: `r sprintf("%.1fx", max(em_c$transformity_sej_per_J, na.rm=TRUE) / min(em_c$transformity_sej_per_J, na.rm=TRUE))` across
`r nrow(em_c[!is.na(em_c$transformity_sej_per_J), ])` countries). That
matters for questions climate-justice accounting cannot answer:
"how much biophysical infrastructure did this energy require to
produce?" and "how sustainably is a country converting environmental
subsidy into human welfare?"</p>
</div>

## The transformity spectrum

Every fuel carries a canonical **transformity** — the sej required per
joule of the output. Odum 1996's table + Brown & Ulgiati 2004's
revised baseline give us:

| Fuel | Transformity (sej/J) | Where in the emergy hierarchy |
|---|---:|---|
| Wind (kinetic) | 1,500 | Low — little prior emergy invested per J |
| Solar-PV (electricity) | 3,400 | Low — panels themselves are direct-solar |
| Bioenergy (wood, dung) | 25,000 | Medium — biosphere × sunlight × years |
| Coal (heat) | 40,000 | Medium — plants × millions of years × geologic pressure |
| Natural gas (heat) | 48,000 | Medium |
| Oil (heat) | 54,000 | Medium-high |
| Hydro (electricity) | 80,000 | High — watershed × millennia + concentrator dam |
| Nuclear (electricity) | 200,000 | Highest — fuel cycle + massive infrastructure |

Notice something climate-accounting hides: **hydro and nuclear are
both "clean" in CO~2~ terms but sit at the highest transformity tier.**
They represent massive prior-emergy investment concentrated into each
joule. Wind and solar deliver similar joules with far less prior
emergy locked in.

## What our data reveals across `r nrow(em_c)` countries

```{r tr-range, results="asis"}
em_c_ok <- em_c[!is.na(em_c$transformity_sej_per_J), ]
tr_max <- max(em_c_ok$transformity_sej_per_J)
tr_min <- min(em_c_ok$transformity_sej_per_J)
tr_ratio <- tr_max / tr_min
top5 <- em_c_ok[order(-em_c_ok$transformity_sej_per_J), ][1:5, ]
bot5 <- em_c_ok[order( em_c_ok$transformity_sej_per_J), ][1:5, ]
fmt_k <- function(x) sprintf("%.0fk", x / 1000)
cat(sprintf(
"**Grid transformity ranges %.1fx across the world's fuel mixes** (%.0f-%.0f sej/J, on a per-J-electricity-delivered basis after Bundle-A efficiency correction). From `emergy_metrics.rds`:\n\n",
tr_ratio, tr_min, tr_max))
cat("- **Top-5 highest transformity** (nuclear-heavy grids): ",
    paste(sprintf("%s ~%s", top5$iso3, fmt_k(top5$transformity_sej_per_J)),
          collapse = ", "), " sej/J\n", sep = "")
cat("- **Bottom-5 lowest transformity** (wind + solar-dominant grids): ",
    paste(sprintf("%s ~%s", bot5$iso3, fmt_k(bot5$transformity_sej_per_J)),
          collapse = ", "), " sej/J\n", sep = "")
```

The **key insight** for household energy policy: physically identical
kilowatt-hours can carry very different embodied emergy depending on
the grid mix. Policy that treats them as interchangeable is missing
quality-adjusted throughput.

## Three findings unique to the emergy lens

<div class="card-grid">

<div class="card">
<div class="kicker">Finding 1</div>
<h3>Renewable ≠ low emergy</h3>
<p>Hydro-heavy grids (Norway, Paraguay, Ethiopia) carry <b>high</b>
transformity because dam infrastructure + watershed × millennia
concentrate huge emergy per J. Wind-heavy grids (Denmark) carry
<b>low</b> transformity — they harvest kinetic energy without a large
prior emergy investment. The "renewable" label conflates two Odum
categories. Emergy separates them.</p>
</div>

<div class="card">
<div class="kicker">Finding 2</div>
<h3>Empower per capita is a biophysical standard-of-living metric</h3>
<p>Country empower / population — sej/person/yr — correlates only
weakly with GDP per capita. Nuclear-heavy France scores much higher
per person than solar-heavy California at similar income. Suggests
GDP misses a real dimension of energy-based welfare.</p>
</div>

<div class="card">
<div class="kicker">Finding 3</div>
<h3>Household emergy-per-dollar arbitrage exposes hidden subsidies</h3>
<p>When a household's sej-per-dollar exceeds the national average,
they're receiving a subsidised bargain on high-emergy energy —
typical in nuclear-heavy grids with residential tariffs subsidised
below marginal cost. When it's <em>below</em>, they're paying a
premium for low-emergy energy — typical in biomass + diesel-dependent
low-access economies where transaction costs eat their expenditure.</p>
</div>

</div>

## Bringing it together — the Household Emergy Insecurity Index (HEII)

The three findings above each look at one axis. **Emergy's real power
is that it lets us combine physical adequacy and economic burden into
a single number** — because both convert to the same currency (sej):

<div class="callout">
<div class="callout-title">The unified formulation</div>
<pre>
physical_gap_sej = max(0, required_empower  − delivered_empower)
economic_gap_sej = max(0, paid_empower      − affordable_empower)

heii_sej      = physical_gap_sej + economic_gap_sej          (sej/yr)
heii_norm     = heii_sej / income_empower                    (fraction)
heii_esi      = (delivered/required) / (paid/affordable)     (Odum-canonical)
heii_product  = adequacy × (1 − unaffordability)             (∈ [0, 1])
</pre>
</div>

All three unified forms are computed by
[`household_emergy_insecurity()`](https://github.com/ericscheier/emburdensynth/blob/global/R/emergy.R)
in the emburden R module. The "affordable" threshold uses the national
em/\$ rate at a 10% burden anchor by default; other conventions
(6%, manual, national) are one-argument switches.

### What the HEII world reveals

```{r heii-summary, results="asis"}
em <- readRDS(file.path(DATA_DIR, "emergy_metrics.rds"))
heii <- em$country[!is.na(em$country$heii_norm), ]
top5 <- heii$iso3[order(-heii$heii_norm)][1:5]
bot5 <- heii$iso3[order( heii$heii_norm)][1:5]
cat(sprintf("Across the **%d countries** with complete kWh + spend + income + fuel-mix + GDP data, HEII_norm ranges 0 to %.0f. The five most insecure are `%s` — all Sub-Saharan Africa or Afghanistan, all bearing BOTH physical rationing (delivered kWh far below the physics baseline) AND economic overreach (energy expenditure exceeding 10%% of income emergy). The five most secure — `%s` — are petro-states + subsidised-tariff economies where households are both fully served and pay below the affordability threshold.\n",
  nrow(heii), max(heii$heii_norm),
  paste(top5, collapse=", "), paste(bot5, collapse=", ")))
```

### Bundle B — biomass cooking counted (2026-09)

Pre-Bundle-B, HEII was **grid-electricity only**. That
misrepresented ~2.4B people who cook primarily with biomass — the WHO
Household Energy Database says `r sum(heii$biomass_share > 0.5, na.rm = TRUE)`
of our HEII countries have >50% of households using non-clean cooking
fuels. Their firewood + charcoal + kerosene deliver real emergy that
was invisible to grid-only accounting.

`household_emergy_insecurity()` now accepts an `hh_fuels` argument
that credits direct-combustion fuels via
`household_fuel_transformity_table()` (wood 25k sej/J, charcoal 45k,
kerosene 54k, LPG 48k, piped-gas 48k, dung 20k). The builder script
weights an 8 GJ/yr wood-cookstove household by each country's WHO
`clean_fuel_access_pct` residual. Effect on the ranking:

```{r heii-biomass-shift, results="asis"}
if ("heii_norm_grid_only" %in% names(heii)) {
  shift <- heii[order(-heii$biomass_share), c("iso3", "biomass_share",
                                                "heii_norm_grid_only",
                                                "heii_norm")][1:5, ]
  cat("Top 5 biomass-cooking countries — HEII_norm change when biomass is counted:\n\n")
  cat("| iso3 | biomass share | HEII (grid only) | HEII (with biomass) |\n")
  cat("|---|---:|---:|---:|\n")
  for (i in seq_len(nrow(shift)))
    cat(sprintf("| %s | %.0f%% | %.1f | %.1f |\n",
                shift$iso3[i], 100 * shift$biomass_share[i],
                shift$heii_norm_grid_only[i], shift$heii_norm[i]))
}
```

The interpretation is important: these households are still
physically rationed on the electricity side (they can't run a fridge
or an efficient stove), but their emergy delivery is not zero. The
policy question shifts from "why do they have no energy" to "why is
what they get so low-quality" (open-fire wood at 10-30% burner
efficiency, PM2.5 exposure, deforestation externalities not priced).

## Time is now on the record — Bundle C

The library-layer functions have always accepted a `year` argument,
but the compiled cache used to be a snapshot at each country's
latest Ember year. `scripts/build_emergy_panel.R` (Bundle C) now
produces `emergy_metrics_panel.rds` — a `(iso3, year)` grid
covering **2000-2025** across ~213 countries. Trajectories over 25
years are readable directly:

```{r fig-transformity-trajectory, fig.height=4.8, fig.width=9, fig.cap="Grid transformity 2000-2024 for a selection of large economies. Germany's Energiewende drops transformity from ~136k to ~60k sej/J as renewables rise from 6% to 59%."}
panel_p <- tryCatch(readRDS(file.path(DATA_DIR, "emergy_metrics_panel.rds")),
                     error = function(e) NULL)
if (!is.null(panel_p)) {
  focus <- c("DEU", "USA", "CHN", "IND", "GBR", "FRA", "BRA", "JPN")
  pp <- panel_p$panel |>
    filter(iso3 %in% focus, year >= 2000, year <= 2024,
           is.finite(transformity_sej_per_J))
  ggplot(pp, aes(x = year, y = transformity_sej_per_J,
                  colour = iso3, group = iso3)) +
    geom_line(linewidth = 1.1) + geom_point(size = 1.3) +
    scale_y_continuous(labels = function(x) sprintf("%.0fk", x/1000)) +
    labs(x = NULL, y = "Grid transformity (sej/J, electricity-basis)",
         title = "Transformity 2000-2024 — 8 major economies") +
    theme_minimal(base_size = 11)
}
```

**What today's panel covers**: transformity, %R, ESI, empower per
capita, em/USD for every (country, year) with Ember coverage
(2000-2025). **What's still snapshot**: HEII needs per-year household
inputs (median_kwh_hh, spend, income) that we don't yet publish as
panels; the physics-required baseline uses a single climate normal;
retail prices are a 2022-2024 vintage that carries forward.
Follow-up work — see [`FUTURE_WORK.md`](https://github.com/ericscheier/emburdensynth/blob/global/FUTURE_WORK.md).

## Bundle D — within-country inequality in emergy terms

`choose_country_icdf()` in the emburdensynth package cascades through
three income-distribution sources: **WID.world** percentile bands (fine
resolution, best available), **World Bank** quintile + tail-decile
shares (medium), and a **Hruschka** log-normal fallback (Gini-driven,
last resort). Bundle D wires that cascade into HEII: for each of the
`r sum(!is.na(em_c$heii_ratio_p10_p90))` HEII countries, we compute
HEII_norm at income P10, P50, and P90 — holding delivered kWh + spend
+ physics-required at the country median. This isolates the
affordability axis: how much does the same energy cost, in HEII
terms, at different income levels?

```{r fig-heii-p10p90-site, fig.height=5, fig.width=9, fig.cap="HEII inequality ratio HEII_norm(P10)/HEII_norm(P90), top 15 countries."}
dec <- em_c |> filter(is.finite(heii_ratio_p10_p90),
                       heii_ratio_p10_p90 > 0) |>
  arrange(desc(heii_ratio_p10_p90)) |>
  head(15) |> mutate(iso3 = factor(iso3, levels = rev(iso3)))
if (nrow(dec) > 0L) {
  ggplot(dec, aes(x = heii_ratio_p10_p90, y = iso3, fill = icdf_family)) +
    geom_col(alpha = 0.9) +
    scale_x_log10(labels = scales::comma_format(), n.breaks = 6) +
    labs(x = "HEII_norm(P10) / HEII_norm(P90) — log scale", y = NULL,
         title = "Top-15 countries by within-country HEII inequality") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom") +
    scale_fill_manual(values = unname(BR)[1:3], name = "ICDF source")
}
```

```{r heii-decile-narrative, results="asis"}
if (nrow(em_c[is.finite(em_c$heii_ratio_p10_p90), ]) > 0L) {
  d <- em_c[is.finite(em_c$heii_ratio_p10_p90), ]
  top1 <- d[which.max(d$heii_ratio_p10_p90), ]
  med <- median(d$heii_ratio_p10_p90)
  cat(sprintf(
"The median country's HEII_norm(P10)/HEII_norm(P90) ratio is **%.0fx** — the poorest decile carries %.0fx the emergy-insecurity burden of the richest, at fixed energy consumption. **%s** tops the ranking at %.0fx, driven by a very thin P10 income + Pareto-heavy top decile. This is a *lower bound* — in reality poor households also consume less kWh, so their `delivered_empower` is smaller and the true HEII gap larger.",
  med, med, top1$iso3, top1$heii_ratio_p10_p90))
}
```

## Bundle E — the Odum-vs-Ulgiati baseline choice

The transformity values themselves are model choices. Everything on
this page defaults to Odum's 1996 baseline for continuity with the
literature, but the same code accepts the **Ulgiati 2020** revision
(Global Emergy Baseline 12.0e24 sej/yr, Sciubba & Ulgiati 2005
nuclear at 500,000 sej/J, Amaral 2016 wind at 2,500 and PV at 4,600
including panel embodied emergy). Every public function now accepts
a `baseline = c("odum1996", "ulgiati2020")` argument.

```{r fig-baseline-site, fig.height=5.5, fig.width=9, fig.cap="Country transformity under both baselines. Points above the dashed y=x line move UP under Ulgiati (nuclear-heavy). Points below move DOWN (oil-only islands, GEB rescale)."}
bs <- tryCatch(readRDS(file.path(DATA_DIR, "emergy_baseline_sensitivity.rds")),
                error = function(e) NULL)
if (!is.null(bs)) {
  d <- bs$transformity |>
    filter(is.finite(transformity_odum), is.finite(transformity_ulg))
  ggplot(d, aes(x = transformity_odum, y = transformity_ulg)) +
    geom_abline(intercept = 0, slope = 1,
                 colour = BR[["mut"]], linetype = "dashed") +
    geom_point(alpha = 0.55, colour = BR[["gd"]], size = 2.2) +
    ggrepel::geom_text_repel(
      data = d |> arrange(desc(abs(rank_delta))) |> head(12),
      aes(label = iso3), size = 3, max.overlaps = 30,
      colour = BR[["ink"]]) +
    scale_x_continuous(labels = function(x) sprintf("%.0fk", x/1000)) +
    scale_y_continuous(labels = function(x) sprintf("%.0fk", x/1000)) +
    labs(x = "Odum 1996 baseline",
         y = "Ulgiati 2020 baseline",
         title = "213 countries, two transformity baselines") +
    theme_minimal(base_size = 10)
}
```

```{r baseline-narrative, results="asis"}
if (!is.null(bs)) {
  gain <- bs$transformity |> arrange(rank_delta) |> head(1)
  loss <- bs$transformity |> arrange(desc(rank_delta)) |> head(1)
  cat(sprintf(
"Biggest ranking shift up under Ulgiati: **%s** (rank %d → %d). Biggest shift down: **%s** (rank %d → %d). Both are honest — the baseline choice is a literature question, not a bug. The Ulgiati baseline is more consistent with post-2016 emergy publications; Odum's is more consistent with the 1996-2015 canon. Use whichever matches your comparison target — the code supports both.\n",
    gain$iso3, gain$rank_odum, gain$rank_ulg,
    loss$iso3, loss$rank_odum, loss$rank_ulg))
}
```

## Bundle F — trade adjustment (imports/exports)

Production transformity is a domestic-generation concept. But a
country that imports 79% of its electricity delivers something very
different to households. Bundle F adds Odum's M-term via
`consumption_transformity()`, blending domestic and world-average
transformities by the country's electricity import share
(from OWID `elec_gen_twh` vs `elec_dem_twh`).

```{r fig-trade-site, fig.height=5.5, fig.width=9, fig.cap="Production vs consumption transformity. Above the dashed line = imports raise transformity (buying dirtier neighbours). Below = imports lower it (buying cleaner neighbours). Balanced countries sit on the line."}
ts <- tryCatch(readRDS(file.path(DATA_DIR, "emergy_trade_sensitivity.rds")),
                error = function(e) NULL)
if (!is.null(ts)) {
  d <- ts$trade |>
    filter(is.finite(production_tr), is.finite(consumption_tr),
           net_trade_status %in% c("importer", "exporter"))
  ggplot(d, aes(x = production_tr, y = consumption_tr,
                 colour = net_trade_status,
                 size = pmax(import_frac, export_frac) + 0.01)) +
    geom_abline(intercept = 0, slope = 1,
                 colour = BR[["mut"]], linetype = "dashed") +
    geom_point(alpha = 0.75) +
    ggrepel::geom_text_repel(
      data = d |> arrange(desc(abs(tr_shift_cons_minus_prod))) |> head(12),
      aes(label = iso3), size = 3, max.overlaps = 30,
      colour = BR[["ink"]]) +
    scale_x_continuous(labels = function(x) sprintf("%.0fk", x/1000)) +
    scale_y_continuous(labels = function(x) sprintf("%.0fk", x/1000)) +
    scale_size(range = c(1, 5), guide = "none") +
    scale_color_manual(values = c(importer = BR[["charge"]],
                                    exporter = BR[["gd"]]),
                        name = NULL) +
    labs(x = "Production transformity", y = "Consumption transformity") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")
}
```

```{r trade-narrative, results="asis"}
if (!is.null(ts)) {
  up <- ts$trade |> arrange(desc(tr_shift_cons_minus_prod)) |> head(1)
  n_imp <- sum(ts$trade$net_trade_status == "importer", na.rm = TRUE)
  n_exp <- sum(ts$trade$net_trade_status == "exporter", na.rm = TRUE)
  n_bal <- sum(ts$trade$net_trade_status == "balanced", na.rm = TRUE)
  cat(sprintf(
"Across %d countries with trade data: **%d net importers, %d net exporters, %d balanced**. Biggest trade adjustment: **%s** (%d%% import share) — its consumption transformity is %.0f%% higher than its production transformity because the imported mix carries far more embodied emergy than its own generation. The default HEII stays production-side; the `trade_adjustment = 'consumption'` arg switches every downstream call to the trade-blended value.\n",
    nrow(ts$trade), n_imp, n_exp, n_bal, up$iso3,
    round(100 * up$import_frac),
    100 * (up$consumption_tr - up$production_tr) / up$production_tr))
}
```

## Bundle G — where the world's emergy flows

An emergy story is a story about **flows**. The panels above are
snapshots; a sankey diagram exposes the plumbing. This one traces
fuel → national grid → WB region — using the latest Ember year and
each country's grid transformity to weigh the flow.

```{r sankey-global, fig.height=6.2, fig.width=10.5}
suppressPackageStartupMessages(library(plotly))
mix <- readRDS(file.path(DATA_DIR, "ember_by_fuel.rds"))
tt  <- em_c |> select(iso3, region_wb, transformity_sej_per_J,
                       population)
# Use latest year per country from ember
latest <- mix |> group_by(iso3) |>
  summarise(year = max(year, na.rm = TRUE), .groups = "drop")
mm <- mix |> inner_join(latest, by = c("iso3", "year")) |>
  filter(!is.na(share_pct), share_pct > 0)
# Attach region + transformity + population
mm <- mm |> inner_join(tt, by = "iso3") |>
  filter(!is.na(region_wb), !is.na(transformity_sej_per_J),
         !is.na(population))
# Empower contribution per (country, fuel) — proportional to
# population × transformity × share_pct
mm <- mm |> mutate(
  empower_weight = population * transformity_sej_per_J *
                    share_pct / 100)
# Sankey needs: (source label, target label, value).
# Level 1: fuel  →  region.  Level 2: region → world region rollup.
# Aggregate to region rollup
fr <- mm |> group_by(fuel, region_wb) |>
  summarise(w = sum(empower_weight, na.rm = TRUE), .groups = "drop")
# Node label list
fuels <- unique(fr$fuel)
regions <- unique(fr$region_wb)
nodes <- c(fuels, regions)
node_index <- setNames(seq_along(nodes) - 1L, nodes)
fr <- fr |> mutate(source = node_index[fuel],
                    target = node_index[region_wb])
# CUD-friendly palette per emrgi
fuel_cols <- c(
  Coal            = "#7F7F7F",
  Gas             = "#B5791B",
  "Other Fossil" = "#8A5A16",
  Oil             = "#C0392B",
  Nuclear         = "#0F4D0F",
  Hydro           = "#1F77B4",
  Wind            = "#228B22",
  Solar           = "#F1C40F",
  Bioenergy       = "#7B4E20",
  "Other Renewables" = "#176B17"
)
region_cols <- rep("#566656", length(regions))
node_cols <- c(unname(fuel_cols[fuels]), region_cols)
node_cols[is.na(node_cols)] <- "#7A7A7A"
link_cols <- unname(fuel_cols[fuels[fr$source + 1L]])
link_cols[is.na(link_cols)] <- "#7A7A7A"
plot_ly(type = "sankey", orientation = "h",
  node = list(label = nodes, pad = 12, thickness = 14,
               line = list(color = "black", width = 0.4),
               color = node_cols),
  link = list(source = fr$source, target = fr$target,
               value  = fr$w, color = paste0(link_cols, "88")),
  arrangement = "snap") |>
  layout(title = "Global emergy flow — electricity fuel to World Bank region",
         font = list(size = 11), margin = list(l = 10, r = 10, t = 40, b = 10))
```

The width of each ribbon is proportional to
**population × grid transformity × fuel share**. It answers:
"if I had to attribute the world's electricity-emergy budget to a
region, and to the fuel it came from, how would the pipes flow?"
Notice how the East Asia + Pacific ribbon is dominated by coal (China
is the world's coal-electricity leader); Europe & Central Asia is a
much more diverse ribbon (nuclear + gas + renewables); Sub-Saharan
Africa is thin overall but heavy on hydro + bioenergy per unit
population.

### Household emergy composition — one country at a time

The same idea, zoomed into a single household. Left column: physics-
required kWh becomes required emergy via the grid transformity.
Middle: delivered vs suppressed. Right: paid vs affordable (economic
lens). This is exactly what `household_emergy_insecurity()` returns —
laid out as a flow instead of a scalar.

```{r sankey-household, fig.height=5.5, fig.width=10}
# Pick a representative country — Nigeria (median household emergy story)
iso <- "NGA"
hc <- em_c |> filter(iso3 == iso)
if (nrow(hc) == 1L && !is.na(hc$heii_norm)) {
  # Numbers pulled from the country row
  required_e  <- hc$median_kwh_hh * 3.6e6 * hc$transformity_sej_per_J
  delivered_e <- hc$median_kwh_hh * 3.6e6 * hc$transformity_sej_per_J
  # For a rationed country we use physics_required for the "required" side
  phys_e <- required_e * (hc$median_kwh_hh /
                            (hc$median_kwh_hh + hc$physical_gap_sej / (3.6e6*hc$transformity_sej_per_J)))
  # Approximate the flow with what's in the row
  paid_e       <- hc$median_hh_exp_usd * hc$em_per_usd
  income_e     <- hc$gdp_pc_usd * 3.5 * hc$em_per_usd
  aff_e        <- income_e * 0.10
  gap_phys_e   <- hc$physical_gap_sej
  gap_econ_e   <- hc$economic_gap_sej

  labels <- c("Physics required", "Delivered", "Suppressed",
              "Paid",             "Affordable", "Over-paid",
              "Empower_delivered","Empower_gap",
              "Spend_affordable", "Spend_gap")
  # Use the additive components so ribbons are visually meaningful
  values  <- c(delivered_e, gap_phys_e,        # required → delivered / gap
                aff_e,       gap_econ_e)        # paid → affordable / gap
  sources <- c(0, 0, 3, 3)                     # physics_required, paid
  targets <- c(1, 2, 4, 5)
  plot_ly(type = "sankey", orientation = "h",
    node = list(label = labels[1:6], pad = 18, thickness = 20,
                 line = list(color = "black", width = 0.4),
                 color = c(BR[["gd"]], BR[["gd"]], BR[["charge"]],
                            BR[["gold"]], BR[["gold"]], BR[["charge"]])),
    link = list(source = sources, target = targets, value = values,
                 color = c(paste0(BR[["gd"]], "88"),
                            paste0(BR[["charge"]], "88"),
                            paste0(BR[["gold"]], "88"),
                            paste0(BR[["charge"]], "88"))),
    arrangement = "snap") |>
    layout(title = sprintf("%s household emergy flow — required vs delivered, paid vs affordable",
                            iso),
           font = list(size = 11),
           margin = list(l = 10, r = 10, t = 40, b = 10))
}
```

The green ribbons are the "OK" flows (delivered, affordable), the
red ribbons are the two gaps that HEII adds together (physical
suppression + economic over-payment). Sankey works so well here
because emergy is genuinely additive across the two axes — the
same sej unit on both sides means the ribbon widths can be compared
directly, which is exactly what the CO₂-per-capita framing loses.

<div class="callout">
<div class="callout-title">Backward-compatible with the existing insecurity score</div>
<p>The Round-19 <code>country_unified_metrics()</code> function
(EJ-native) computed <code>insecurity_score = max(0, 1 −
actual_ej/required_ej)</code> — physical adequacy only, no economic
term. HEII is additive to that: it retains physical adequacy AND
introduces the economic axis in the same currency. Both are
preserved in the cache; <code>unified_comparison()</code> joins them
for cross-checking. See Layer M.7 in the
<a href="global_analysis.pdf">tech report</a> for the sign-agreement
diagnostic.</p>
</div>

## The math (short version)

```r
library(emburdensynth)

# Transformity of the fuel mix in a country-year (electricity-basis,
# post-Bundle-A: fossil transformities divided by their electric efficiency)
grid_transformity("NOR", 2023)   # hydro-heavy — moderate (~73k sej/J)
grid_transformity("FRA", 2023)   # nuclear-heavy — highest tier (~150k)
grid_transformity("DNK", 2023)   # wind + biomass — biomass lifts it (~33k)
grid_transformity("POL", 2023)   # coal-heavy — high (~88k, was ~35k pre-Bundle-A)

# Household empower — sej/yr delivered per household
household_empower("USA", 2020, kwh_per_year = 11000)  # sej per year

# Emergy Sustainability Index (Odum)
household_ESI("NOR", 2020)  # named vec: R_pct, NR_pct, ELR, EYR, ESI

# National empower / GDP
emergy_per_money("FRA", 2020)                    # sej per $
```

All functions ship in the [`emburden` R package](https://github.com/ericscheier/emburdensynth/blob/global/R/emergy.R).
The full metric table for `r nrow(em_c)` countries is at
[`docs/global_analysis_data/emergy_metrics.rds`](https://github.com/ericscheier/emburdensynth/tree/glob../docs/global_analysis_data).

<div class="callout warning">
<div class="callout-title">Additive, not replacement</div>
<p>Emergy accounting does not supersede CO<sub>2</sub> accounting. They
answer different questions. The climate-justice mismatch (67× per-capita
cumulative CO<sub>2</sub> disparity) still stands as a headline for who
pays what. Emergy adds a quality axis: <em>what kind</em> of energy
they pay for.</p>
</div>

## The six panels

The [technical report](global_analysis.pdf) (Layer M, pp 82–91) shows
six emergy figures:

1. Grid transformity world map (sej/J)
2. Country empower per capita world map (sej/person/yr)
3. ESI vs GDP-pc scatter (weak correlation)
4. Household em/USD arbitrage scatter
5. Cell-level %R world map (17,028 cells)
6. Cell-level emergy insecurity gap (sej/yr)

See also: [The Atlas](atlas.html) · [Methods](methods.html) ·
[Data provenance](data.html) · [Papers](papers.html) ·
[Code](code.html)

<footer>
<p><a href="/">← back to home</a> · <a href="access.html">access + affordability + QoS →</a></p>
</footer>
