---
title: "Emergy — Odum's quality-of-energy lens"
pagetitle: "emburden.org — Emergy"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.align="center", out.width="100%",
                      fig.width=10, fig.height=6, dpi=110)
suppressPackageStartupMessages({library(dplyr); library(ggplot2); library(sf)})
DATA_DIR <- "../docs/global_analysis_data"
# Bundled brand palette (matches emburdensynR/brand_palette.R:emrgi_brand_colors)
BR <- c(g = "#228b22", gd = "#176b17", gdd = "#0f4d0f", gold = "#b5791b",
        ink = "#1d2a1d", mut = "#566656", line = "#d6e4d6", bg = "#f3f8f3",
        charge = "#c0392b", warn = "#8a5a16")
source("R/counts.R")
em <- readRDS(file.path(DATA_DIR, "emergy_metrics.rds"))
em_c <- em$country
em_cell <- em$cell
world <- tryCatch(readRDS(file.path(DATA_DIR, "world_sf.rds")), error=function(e) NULL)
```

# Emergy — a different currency for energy analysis

Climate-justice accounting uses **CO~2~ emissions** as the currency.
Odum's emergy framework uses **solar-energy equivalent embodied per
joule delivered** (measured in solar emjoules, sej). Same physical
kilowatt-hour at the socket can carry wildly different prior emergy
investment depending on the fuel that generated it.

<div class="callout">
<div class="callout-title">Why care about a second currency</div>
<p>Two grids that deliver identical kWh per capita can differ
<b>15-fold</b> in the emergy they concentrate into each joule. That
matters for questions climate-justice accounting cannot answer:
"how much biophysical infrastructure did this energy require to
produce?" and "how sustainably is a country converting environmental
subsidy into human welfare?"</p>
</div>

## The transformity spectrum

Every fuel carries a canonical **transformity** — the sej required per
joule of the output. Odum 1996's table + Brown & Ulgiati 2004's
revised baseline give us:

| Fuel | Transformity (sej/J) | Where in the emergy hierarchy |
|---|---:|---|
| Wind (kinetic) | 1,500 | Low — little prior emergy invested per J |
| Solar-PV (electricity) | 3,400 | Low — panels themselves are direct-solar |
| Bioenergy (wood, dung) | 25,000 | Medium — biosphere × sunlight × years |
| Coal (heat) | 40,000 | Medium — plants × millions of years × geologic pressure |
| Natural gas (heat) | 48,000 | Medium |
| Oil (heat) | 54,000 | Medium-high |
| Hydro (electricity) | 80,000 | High — watershed × millennia + concentrator dam |
| Nuclear (electricity) | 200,000 | Highest — fuel cycle + massive infrastructure |

Notice something climate-accounting hides: **hydro and nuclear are
both "clean" in CO~2~ terms but sit at the highest transformity tier.**
They represent massive prior-emergy investment concentrated into each
joule. Wind and solar deliver similar joules with far less prior
emergy locked in.

## What our data reveals across `r nrow(em_c)` countries

**Grid transformity ranges 15× across the world's fuel mixes.** From
`emergy_metrics.rds`:

- **Top-5 highest transformity**: nuclear-heavy grids
  (France ~150k, Slovakia ~145k, Ukraine ~140k, Slovenia ~125k,
  Czech Republic ~115k sej/J)
- **Bottom-5 lowest transformity**: wind + biomass-heavy grids
  (Denmark ~10k, Germany ~15k, Luxembourg ~13k, Lithuania ~18k,
  Estonia ~20k sej/J)

The **key insight** for household energy policy: a Danish
kilowatt-hour and a French kilowatt-hour are physically identical but
Odum-different by an order of magnitude. Policy that treats them as
interchangeable is missing quality-adjusted throughput.

## Three findings unique to the emergy lens

<div class="card-grid">

<div class="card">
<div class="kicker">Finding 1</div>
<h3>Renewable ≠ low emergy</h3>
<p>Hydro-heavy grids (Norway, Paraguay, Ethiopia) carry <b>high</b>
transformity because dam infrastructure + watershed × millennia
concentrate huge emergy per J. Wind-heavy grids (Denmark) carry
<b>low</b> transformity — they harvest kinetic energy without a large
prior emergy investment. The "renewable" label conflates two Odum
categories. Emergy separates them.</p>
</div>

<div class="card">
<div class="kicker">Finding 2</div>
<h3>Empower per capita is a biophysical standard-of-living metric</h3>
<p>Country empower / population — sej/person/yr — correlates only
weakly with GDP per capita. Nuclear-heavy France scores much higher
per person than solar-heavy California at similar income. Suggests
GDP misses a real dimension of energy-based welfare.</p>
</div>

<div class="card">
<div class="kicker">Finding 3</div>
<h3>Household emergy-per-dollar arbitrage exposes hidden subsidies</h3>
<p>When a household's sej-per-dollar exceeds the national average,
they're receiving a subsidised bargain on high-emergy energy —
typical in nuclear-heavy grids with residential tariffs subsidised
below marginal cost. When it's <em>below</em>, they're paying a
premium for low-emergy energy — typical in biomass + diesel-dependent
low-access economies where transaction costs eat their expenditure.</p>
</div>

</div>

## Bringing it together — the Household Emergy Insecurity Index (HEII)

The three findings above each look at one axis. **Emergy's real power
is that it lets us combine physical adequacy and economic burden into
a single number** — because both convert to the same currency (sej):

<div class="callout">
<div class="callout-title">The unified formulation</div>
<pre>
physical_gap_sej = max(0, required_empower  − delivered_empower)
economic_gap_sej = max(0, paid_empower      − affordable_empower)

heii_sej      = physical_gap_sej + economic_gap_sej          (sej/yr)
heii_norm     = heii_sej / income_empower                    (fraction)
heii_esi      = (delivered/required) / (paid/affordable)     (Odum-canonical)
heii_product  = adequacy × (1 − unaffordability)             (∈ [0, 1])
</pre>
</div>

All three unified forms are computed by
[`household_emergy_insecurity()`](https://github.com/ericscheier/emburdensynth/blob/global/R/emergy.R)
in the emburden R module. The "affordable" threshold uses the national
em/\$ rate at a 10% burden anchor by default; other conventions
(6%, manual, national) are one-argument switches.

### What the HEII world reveals

```{r heii-summary, results="asis"}
em <- readRDS(file.path(DATA_DIR, "emergy_metrics.rds"))
heii <- em$country[!is.na(em$country$heii_norm), ]
top5 <- heii$iso3[order(-heii$heii_norm)][1:5]
bot5 <- heii$iso3[order( heii$heii_norm)][1:5]
cat(sprintf("Across the **%d countries** with complete kWh + spend + income + fuel-mix + GDP data, HEII_norm ranges 0 to %.0f. The five most insecure are `%s` — all Sub-Saharan Africa or Afghanistan, all bearing BOTH physical rationing (delivered kWh far below the physics baseline) AND economic overreach (energy expenditure exceeding 10%% of income emergy). The five most secure — `%s` — are petro-states + subsidised-tariff economies where households are both fully served and pay below the affordability threshold.\n",
  nrow(heii), max(heii$heii_norm),
  paste(top5, collapse=", "), paste(bot5, collapse=", ")))
```

<div class="callout">
<div class="callout-title">Backward-compatible with the existing insecurity score</div>
<p>The Round-19 <code>country_unified_metrics()</code> function
(EJ-native) computed <code>insecurity_score = max(0, 1 −
actual_ej/required_ej)</code> — physical adequacy only, no economic
term. HEII is additive to that: it retains physical adequacy AND
introduces the economic axis in the same currency. Both are
preserved in the cache; <code>unified_comparison()</code> joins them
for cross-checking. See Layer M.7 in the
<a href="global_analysis.pdf">tech report</a> for the sign-agreement
diagnostic.</p>
</div>

## The math (short version)

```r
library(emburdensynth)

# Transformity of the fuel mix in a country-year
grid_transformity("NOR", 2020)                   # ~74k sej/J (hydro-heavy)
grid_transformity("FRA", 2020)                   # ~150k sej/J (nuclear-heavy)
grid_transformity("DNK", 2020)                   # ~10k sej/J (wind-heavy)

# Household empower — sej/yr delivered per household
household_empower("USA", 2020, kwh_per_year = 11000)  # sej per year

# Emergy Sustainability Index (Odum)
household_ESI("NOR", 2020)  # named vec: R_pct, NR_pct, ELR, EYR, ESI

# National empower / GDP
emergy_per_money("FRA", 2020)                    # sej per $
```

All functions ship in the [`emburden` R package](https://github.com/ericscheier/emburdensynth/blob/global/R/emergy.R).
The full metric table for `r nrow(em_c)` countries is at
[`docs/global_analysis_data/emergy_metrics.rds`](https://github.com/ericscheier/emburdensynth/tree/glob../docs/global_analysis_data).

<div class="callout warning">
<div class="callout-title">Additive, not replacement</div>
<p>Emergy accounting does not supersede CO<sub>2</sub> accounting. They
answer different questions. The climate-justice mismatch (67× per-capita
cumulative CO<sub>2</sub> disparity) still stands as a headline for who
pays what. Emergy adds a quality axis: <em>what kind</em> of energy
they pay for.</p>
</div>

## The six panels

The [technical report](global_analysis.pdf) (Layer M, pp 82–91) shows
six emergy figures:

1. Grid transformity world map (sej/J)
2. Country empower per capita world map (sej/person/yr)
3. ESI vs GDP-pc scatter (weak correlation)
4. Household em/USD arbitrage scatter
5. Cell-level %R world map (17,028 cells)
6. Cell-level emergy insecurity gap (sej/yr)

See also: [The Atlas](atlas.html) · [Methods](methods.html) ·
[Data provenance](data.html) · [Papers](papers.html) ·
[Code](code.html)

<footer>
<p><a href="/">← back to home</a> · <a href="access.html">access + affordability + QoS →</a></p>
</footer>
